---
title: "Review California Bombus Candidate Host Records"
author: "Gretchen"
format: html
execute:
  echo: true
  warning: false
  message: true
editor: 
  markdown: 
    wrap: 72
---

## Overview

This script reviews candidate host-plant records extracted from GBIF
California *Bombus* occurrence records.

The previous script used a broad text search to find possible host,
plant, floral, pollen, nectar, or related associations. Those records
are candidates only. They need manual review before they are treated as
confirmed bee-plant associations.

This script does four main things:

1.  Reads the candidate host-record file.
2.  Creates or updates a manual review CSV.
3.  Reads the reviewed CSV and keeps confirmed records.
4.  Summarizes confirmed records by bee species, host plant type, and
    county.

## Review workflow

This script is designed to be run more than once.

First, run the script to create the manual review file:

`outputs/gbif_bombus_california_host_review.csv`

Then open that CSV in a spreadsheet editor and review the candidate
records by hand. Each row is one candidate bee-plant association.

After editing and saving the review CSV, rerun this script. The script
will preserve the manual review columns and create confirmed-record
tables and summary tables.

The basic review sequence is:

1.  Run this script once.
2.  Open `outputs/gbif_bombus_california_host_review.csv`.
3.  Inspect the candidate evidence in `host_text`.
4.  Mark each row as confirmed, rejected, uncertain, or still needing
    review.
5.  Clean the plant name or host category for confirmed rows.
6.  Save the review CSV with the same filename.
7.  Rerun this script to create summaries.

The summary tables use the `host_plant_type` column. That column should
contain the plant grouping you want to count by. It may be a plant
species, genus, family, or broader category, depending on the quality of
the evidence.

## Setup

```{r setup}
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(tibble)
library(purrr)

# Vectorized helper for testing whether character values contain text.
# This works inside mutate(), filter(), and case_when().
has_text <- function(x) {
  !is.na(x) & nzchar(x)
}

# Add missing columns to a data frame.
#
# GBIF-derived files may differ in which fields are present. This helper
# keeps the script from failing when an expected column is absent.
add_missing_cols <- function(dat, cols) {
  missing_cols <- setdiff(cols, names(dat))

  for (col in missing_cols) {
    dat[[col]] <- NA_character_
  }

  dat
}

# Find the project root by walking upward from the current Quarto file.
#
# The project root is identified by the presence of scripts/, data/,
# and outputs/.
find_project_root <- function() {
  input_file <- tryCatch(knitr::current_input(), error = function(e) NULL)

  if (length(input_file) == 1 &&
      !is.na(input_file) &&
      nzchar(input_file) &&
      file.exists(input_file)) {
    start_dir <- dirname(normalizePath(
      input_file,
      winslash = "/",
      mustWork = TRUE
    ))
  } else {
    start_dir <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
  }

  current_dir <- start_dir

  for (i in seq_len(10)) {
    has_project_dirs <- all(dir.exists(file.path(
      current_dir,
      c("scripts", "data", "outputs")
    )))

    if (has_project_dirs) {
      return(current_dir)
    }

    parent_dir <- dirname(current_dir)

    if (identical(parent_dir, current_dir)) {
      break
    }

    current_dir <- parent_dir
  }

  stop(
    "Could not find project root. Expected scripts/, data/, and outputs/."
  )
}

project_root <- find_project_root()

# Small wrapper to build paths relative to the project root.
path_from_root <- function(...) {
  file.path(project_root, ...)
}

outputs_dir <- path_from_root("outputs")

# Put summary tables in their own output subfolder.
summary_dir <- path_from_root("outputs", "host_review_tables")

dir.create(outputs_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(summary_dir, showWarnings = FALSE, recursive = TRUE)

# Input from the previous GBIF host-record extraction script.
candidate_csv <- path_from_root(
  "outputs",
  "gbif_bombus_california_candidate_host_records.csv"
)

# This is the file you edit by hand.
review_csv <- path_from_root(
  "outputs",
  "gbif_bombus_california_host_review.csv"
)

# These files are derived from the reviewed file.
confirmed_csv <- path_from_root(
  "outputs",
  "gbif_bombus_california_confirmed_host_records.csv"
)

review_issues_csv <- path_from_root(
  "outputs",
  "gbif_bombus_california_host_review_issues.csv"
)

summary_long_csv <- file.path(
  summary_dir,
  "summary_species_host_type_county_long.csv"
)

summary_county_wide_csv <- file.path(
  summary_dir,
  "summary_species_host_type_by_county_wide.csv"
)

summary_host_wide_csv <- file.path(
  summary_dir,
  "summary_species_county_by_host_type_wide.csv"
)

message("Project root: ", project_root)
message("Candidate records: ", candidate_csv)
message("Review file: ", review_csv)
```

## Read candidate host records

```{r read-candidate-records}
if (!file.exists(candidate_csv)) {
  stop("Candidate host-record file not found: ", candidate_csv)
}

candidate_records <- readr::read_csv(
  candidate_csv,
  col_types = readr::cols(.default = readr::col_character()),
  show_col_types = FALSE
)

# These columns are expected from the previous workflow, but some may be
# absent if the source file had limited information.
candidate_records <- add_missing_cols(
  candidate_records,
  c(
    "scientificName",
    "family",
    "genus",
    "specificEpithet",
    "eventDate",
    "county",
    "locality",
    "decimalLatitude",
    "decimalLongitude",
    "source_file",
    "source_row",
    "host_field",
    "host_text"
  )
)

# Create convenient review fields.
#
# bee_species is built in order of preference:
# 1. scientificName, if available
# 2. genus + specificEpithet, if available
# 3. genus alone
# 4. missing
candidate_records <- candidate_records %>%
  mutate(
    bee_species = case_when(
      has_text(scientificName) ~ scientificName,
      has_text(genus) & has_text(specificEpithet) ~
        paste(genus, specificEpithet),
      has_text(genus) ~ genus,
      TRUE ~ NA_character_
    ),
    county = if_else(has_text(county), county, NA_character_)
  )

# Quick diagnostic: where did the candidate text come from?
candidate_records %>%
  count(source_file, host_field, sort = TRUE)
```

## Create or update review file

```{r create-review-file}
# These are the columns intended for manual review.
#
# Do not remove these from the CSV if you edit it by hand. The script
# expects to find them on later runs.
review_columns <- c(
  "review_status",
  "confirmed_host",
  "host_plant_name",
  "host_plant_type",
  "plant_family",
  "plant_genus",
  "plant_species",
  "association_type",
  "reviewer_notes"
)

# These columns identify a candidate record.
#
# They are used to join previous manual review decisions back onto the
# current candidate table. This lets the candidate file be regenerated
# without losing previous review work.
review_keys <- c(
  "source_file",
  "source_row",
  "host_field",
  "host_text"
)

front_review_columns <- c(
  "review_id",
  "host_field",
  "host_text",
  review_columns
)

review_template <- candidate_records %>%
  mutate(
    review_id = as.character(row_number()),
    review_status = "needs_review",
    confirmed_host = "",
    host_plant_name = "",
    host_plant_type = "",
    plant_family = "",
    plant_genus = "",
    plant_species = "",
    association_type = "",
    reviewer_notes = ""
  ) %>%
  relocate(
    any_of(front_review_columns),
    .before = source_file
  )

# If a review file already exists, preserve the manual review fields.
#
# This is the key protection against accidentally overwriting review
# decisions during reruns.
if (file.exists(review_csv)) {
  existing_review <- readr::read_csv(
    review_csv,
    col_types = readr::cols(.default = readr::col_character()),
    show_col_types = FALSE
  )

  existing_review <- add_missing_cols(existing_review, review_columns)
  existing_review <- add_missing_cols(existing_review, review_keys)

  existing_review_manual <- existing_review %>%
    select(any_of(c(review_keys, review_columns))) %>%
    distinct(across(any_of(review_keys)), .keep_all = TRUE)

  review_template <- review_template %>%
    select(-any_of(review_columns)) %>%
    left_join(existing_review_manual, by = review_keys) %>%
    mutate(
      review_status = if_else(
        has_text(review_status),
        review_status,
        "needs_review"
      ),
      across(
        any_of(setdiff(review_columns, "review_status")),
        ~ if_else(has_text(.x), .x, "")
      )
    ) %>%
   relocate(any_of(front_review_columns), .before = source_file)
}

readr::write_csv(review_template, review_csv, na = "")

message("Review file written to: ", review_csv)

# This lets you see how much review has been completed.
review_template %>%
  count(review_status, confirmed_host, sort = TRUE)
```

## Review instructions

Edit this file by hand:

`outputs/gbif_bombus_california_host_review.csv`

The most important evidence column is `host_text`. That is the text that
triggered the candidate host-plant search.

Useful columns to inspect during review are:

```{r review-evidence-columns}
review_evidence_columns <- tibble::tribble(
  ~column, ~what_to_check,

  "bee_species",
  "Bumble bee taxon associated with the candidate record.",

  "county",
  "County where the specimen or observation was recorded.",

  "source_file",
  "Whether the text came from occurrence, verbatim, or extension data.",

  "host_field",
  "The GBIF or Darwin Core field where candidate text was found.",

  "host_text",
  "The original candidate text. This is usually the key review field."
)

knitr::kable(review_evidence_columns)
```

Use `confirmed` and `yes` only when the text gives a usable bee-plant
association.

Use `reject` and `no` when the text is not a usable host or floral
record.

Use `uncertain` when the text suggests a possible association but is
ambiguous.

```{r review-decision-guide}
review_decision_guide <- tibble::tribble(
  ~column, ~suggested_values, ~meaning,

  "review_status",
  "needs_review; confirmed; reject; uncertain",
  "Overall review decision for the candidate record.",

  "confirmed_host",
  "yes; no; uncertain",
  "Whether the text is a usable bee-plant association.",

  "host_plant_name",
  "free text",
  "Cleaned plant name as you want it reported.",

  "host_plant_type",
  "free text",
  "Plant grouping used in summary tables.",

  "plant_family",
  "free text",
  "Plant family, if known.",

  "plant_genus",
  "free text",
  "Plant genus, if known.",

  "plant_species",
  "free text",
  "Plant species, if known.",

  "association_type",
  "floral_visit; pollen; nectar; collected_on; habitat; other",
  "Kind of bee-plant association inferred from the text.",

  "reviewer_notes",
  "free text",
  "Notes about uncertainty, interpretation, or cleanup."
)

knitr::kable(review_decision_guide)
```

Minimum entries for a confirmed row:

```{r confirmed-row-guide}
confirmed_row_guide <- tibble::tribble(
  ~column, ~value,

  "review_status",
  "confirmed",

  "confirmed_host",
  "yes",

  "host_plant_type",
  "Plant category to use in summaries.",

  "host_plant_name",
  "Cleaned plant name, if available."
)

knitr::kable(confirmed_row_guide)
```

Minimum entries for a rejected row:

```{r rejected-row-guide}
rejected_row_guide <- tibble::tribble(
  ~column, ~value,

  "review_status",
  "reject",

  "confirmed_host",
  "no"
)

knitr::kable(rejected_row_guide)
```

## After manual review

After editing the review CSV, save it with the same filename:

`outputs/gbif_bombus_california_host_review.csv`

Then rerun this script.

The chunks below use only rows marked as confirmed. A row is treated as
confirmed when it has a positive review decision and no negative review
decision.

Positive review values are:

- `confirmed_host` is `yes`, `y`, or `true`
- `review_status` is `confirmed`, `accept`, or `accepted`

Negative review values are:

- `confirmed_host` is `no`, `n`, or `false`
- `review_status` is `reject` or `rejected`

If a row has both positive and negative values, it is flagged as a
review conflict and is not used in the final summaries.

## Read reviewed records

```{r read-reviewed-records}
reviewed_records <- readr::read_csv(
  review_csv,
  col_types = readr::cols(.default = readr::col_character()),
  show_col_types = FALSE
)

reviewed_records <- add_missing_cols(reviewed_records, review_columns)

reviewed_records <- add_missing_cols(
  reviewed_records,
  c(
    "bee_species",
    "scientificName",
    "genus",
    "specificEpithet",
    "county",
    "source_file",
    "source_row",
    "host_field",
    "host_text"
  )
)

# Normalize review fields.
#
# str_squish() trims leading/trailing whitespace and collapses repeated
# internal spaces. This helps avoid accidental categories like
# "confirmed " and "confirmed" being treated as different values.
reviewed_records <- reviewed_records %>%
  mutate(
    review_status = str_to_lower(str_squish(review_status)),
    confirmed_host = str_to_lower(str_squish(confirmed_host)),
    host_plant_name = str_squish(host_plant_name),
    host_plant_type = str_squish(host_plant_type),
    plant_family = str_squish(plant_family),
    plant_genus = str_squish(plant_genus),
    plant_species = str_squish(plant_species),
    association_type = str_squish(association_type),
    reviewer_notes = str_squish(reviewer_notes),
    positive_review = confirmed_host %in% c("yes", "y", "true") |
      review_status %in% c("confirmed", "accept", "accepted"),
    negative_review = confirmed_host %in% c("no", "n", "false") |
      review_status %in% c("reject", "rejected"),
    review_conflict = positive_review & negative_review,
    is_confirmed_host = positive_review & !negative_review
  )

reviewed_records %>%
  count(
    review_status,
    confirmed_host,
    is_confirmed_host,
    review_conflict,
    sort = TRUE
  )
```

## Review issues

```{r review-issues}
review_issues <- reviewed_records %>%
  filter(
    review_conflict |
      (is_confirmed_host & !has_text(host_plant_type) &
         !has_text(host_plant_name) & !has_text(plant_species) &
         !has_text(plant_genus) & !has_text(plant_family))
  )

readr::write_csv(review_issues, review_issues_csv, na = "")

message("Review issues written to: ", review_issues_csv)

review_issues %>%
  select(
    review_id,
    review_status,
    confirmed_host,
    bee_species,
    county,
    host_plant_type,
    host_plant_name,
    host_text,
    reviewer_notes
  ) %>%
  head(25)
```

## Confirmed host records

```{r confirmed-host-records}
confirmed_host_records <- reviewed_records %>%
  filter(is_confirmed_host, !review_conflict) %>%
  mutate(
    # Rebuild bee_species if it was missing from the review file.
    bee_species = case_when(
      has_text(bee_species) ~ bee_species,
      has_text(scientificName) ~ scientificName,
      has_text(genus) & has_text(specificEpithet) ~
        paste(genus, specificEpithet),
      has_text(genus) ~ genus,
      TRUE ~ "unknown bee"
    ),

    county = if_else(has_text(county), county, "unknown county"),

    # host_plant_type is the grouping used in the summary tables.
    #
    # This lets you summarize at whatever level is appropriate:
    # species, genus, family, or a broader plant category.
    host_plant_type = case_when(
      has_text(host_plant_type) ~ host_plant_type,
      has_text(host_plant_name) ~ host_plant_name,
      has_text(plant_species) ~ plant_species,
      has_text(plant_genus) ~ plant_genus,
      has_text(plant_family) ~ plant_family,
      TRUE ~ "unassigned host"
    )
  )

readr::write_csv(confirmed_host_records, confirmed_csv, na = "")

message("Confirmed host records written to: ", confirmed_csv)

confirmed_host_records %>%
  select(
    bee_species,
    county,
    host_plant_type,
    host_plant_name,
    association_type,
    source_file,
    host_field,
    host_text
  ) %>%
  head(25)
```

## Summary-table logic

The summary tables count confirmed records by:

- bumble bee species
- host plant type
- county

The `host_plant_type` field controls the plant grouping used in these
tables. If `host_plant_type` is blank, the script falls back to
`host_plant_name`, then `plant_species`, then `plant_genus`, then
`plant_family`.

This allows summaries at different taxonomic levels, depending on how
precise the original text is.

## Summary: species by host plant type by county

```{r summary-species-host-county-long}
# Long format is usually the safest analysis format.
# Each row is one bee species x host type x county combination.
summary_species_host_county <- confirmed_host_records %>%
  count(
    bee_species,
    host_plant_type,
    county,
    name = "n_records"
  ) %>%
  arrange(
    bee_species,
    host_plant_type,
    county
  )

readr::write_csv(summary_species_host_county, summary_long_csv, na = "")

message("Long summary written to: ", summary_long_csv)

knitr::kable(
  summary_species_host_county,
  caption = "Confirmed host records by bee species, host plant type, and county."
)
```

## Wide table: counties as columns

```{r summary-counties-wide}
# This format is useful for quick inspection.
#
# Rows are bee species x host plant type.
# Columns are counties.
summary_species_host_by_county_wide <- summary_species_host_county %>%
  tidyr::pivot_wider(
    names_from = county,
    values_from = n_records,
    values_fill = 0
  ) %>%
  arrange(bee_species, host_plant_type)

readr::write_csv(
  summary_species_host_by_county_wide,
  summary_county_wide_csv,
  na = ""
)

message("County-wide summary written to: ", summary_county_wide_csv)

knitr::kable(
  summary_species_host_by_county_wide,
  caption = paste(
    "Confirmed host records by bee species and host plant type,",
    "with counties as columns."
  )
)
```

## Wide table: host plant types as columns

```{r summary-host-types-wide}
# This format is useful if you want to compare host categories within
# each bee species and county.
#
# Rows are bee species x county.
# Columns are host plant types.
summary_species_county_by_host_wide <- summary_species_host_county %>%
  tidyr::pivot_wider(
    names_from = host_plant_type,
    values_from = n_records,
    values_fill = 0
  ) %>%
  arrange(bee_species, county)

readr::write_csv(
  summary_species_county_by_host_wide,
  summary_host_wide_csv,
  na = ""
)

message("Host-type-wide summary written to: ", summary_host_wide_csv)

knitr::kable(
  summary_species_county_by_host_wide,
  caption = paste(
    "Confirmed host records by bee species and county,",
    "with host plant types as columns."
  )
)
```

## Output files

```{r output-files}
output_files <- tibble::tribble(
  ~file, ~description,

  review_csv,
  "Manual review file. Edit this to confirm or reject records.",

  confirmed_csv,
  "Confirmed host records after manual review.",

  review_issues_csv,
  "Rows with review conflicts or missing host categories.",

  summary_long_csv,
  "Long-format counts by bee species, host plant type, and county.",

  summary_county_wide_csv,
  "Wide-format counts with counties as columns.",

  summary_host_wide_csv,
  "Wide-format counts with host plant types as columns."
)

knitr::kable(output_files)
```
