---
title: "Crotch's Bumble Bee Host Plant Analysis"
author: "OpenAI"
format:
  html:
    toc: true
    code-fold: false
execute:
  echo: true
  warning: true
  message: true
editor: visual
---

## Overview

This Quarto document reproduces the interactive R console workflow as a script. It:

- reads the bee observation data from `../data/`
- converts the table to an `sf` object using longitude and latitude
- downloads California county boundaries with `tigris`
- extracts Santa Barbara County observations
- summarizes host plant observations
- writes figures and tables to `../output/`

## Project layout

Expected project structure:

```text
project/
├── data/
│   └── Crotchii_xerces_1_dec_2025.xlsx
├── output/
└── scripts/
    └── bee_host_plants.qmd
```

## Packages

```{r}
#| label: setup
#| include: true

required_packages <- c(
  "sf",
  "dplyr",
  "ggplot2",
  "tigris",
  "readxl",
  "stringr",
  "readr"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

library(sf)
library(dplyr)
library(ggplot2)
library(tigris)
library(readxl)
library(stringr)
library(readr)

options(tigris_use_cache = TRUE)
```

If you need to reinstall macOS binary packages manually, these were downloaded during your console session:

```text
/var/folders/pn/d57bb9jj719fxgk9pqfgqtlw0000gn/T//RtmpJ5lLOB/downloaded_packages
```

That temporary folder is not reliable long-term, so this script installs from CRAN when needed.

## Paths

```{r}
#| label: paths

script_dir <- getwd()
data_dir <- normalizePath(file.path(script_dir, "..", "data"), mustWork = FALSE)
output_dir <- normalizePath(file.path(script_dir, "..", "output"), mustWork = FALSE)

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

xlsx_path <- file.path(data_dir, "Crotchii_xerces_1_dec_2025.xlsx")
csv_path  <- file.path(data_dir, "Crotchii_xerces_1_dec_2025.csv")
```

## Import data

This section tries an Excel file first, then a CSV file.

```{r}
#| label: import-data

if (file.exists(xlsx_path)) {
  bee_data <- read_excel(xlsx_path, sheet = 1)
} else if (file.exists(csv_path)) {
  bee_data <- read_csv(csv_path, show_col_types = FALSE)
} else {
  stop(
    "No input data file found. Place either 'Crotchii_xerces_1_dec_2025.xlsx' or ",
    "'Crotchii_xerces_1_dec_2025.csv' in ../data/."
  )
}

bee_data <- as.data.frame(bee_data)

glimpse(bee_data)
```

## Convert to spatial data

```{r}
#| label: make-sf

required_columns <- c("decimalLongitude", "decimalLatitude", "Plant_species")
missing_columns <- setdiff(required_columns, names(bee_data))

if (length(missing_columns) > 0) {
  stop("Missing required columns: ", paste(missing_columns, collapse = ", "))
}

bee_sf <- bee_data %>%
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude)) %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

bee_sf_clean <- bee_sf %>%
  filter(!is.na(Plant_species), Plant_species != "")

print(class(bee_sf_clean))
print(st_geometry_type(bee_sf_clean))
```

## County boundaries

```{r}
#| label: california-counties

ca_counties <- counties(state = "CA", cb = TRUE, class = "sf") %>%
  st_transform(st_crs(bee_sf_clean))

santa_barbara <- ca_counties %>%
  filter(NAME == "Santa Barbara") %>%
  st_transform(st_crs(bee_sf_clean))
```

## Santa Barbara County observations

```{r}
#| label: santa-barbara-join

sb_bees <- st_join(bee_sf_clean, santa_barbara, join = st_within, left = FALSE)

sb_plant_counts <- sb_bees %>%
  st_drop_geometry() %>%
  filter(!is.na(Plant_species), Plant_species != "") %>%
  count(Plant_species, sort = TRUE)

top_sb_plants <- sb_plant_counts %>%
  slice_head(n = 10)

top_sb_plants
```

```{r}
#| label: save-santa-barbara-table

write_csv(top_sb_plants, file.path(output_dir, "top_santa_barbara_plants.csv"))
```

```{r}
#| label: santa-barbara-plot

p_sb <- ggplot(top_sb_plants, aes(x = reorder(Plant_species, n), y = n)) +
  geom_col(fill = "darkolivegreen3") +
  coord_flip() +
  labs(
    title = "Host Plant Preferences in Santa Barbara County",
    x = "Plant Species",
    y = "Number of Observations"
  ) +
  theme_minimal()

p_sb

ggsave(
  filename = file.path(output_dir, "host_plant_preferences_santa_barbara.png"),
  plot = p_sb,
  width = 8,
  height = 6,
  dpi = 300
)
```

## Statewide map of host plant observations

```{r}
#| label: statewide-map

p_ca <- ggplot() +
  geom_sf(data = ca_counties, fill = "gray95", color = "gray50", linewidth = 0.2) +
  geom_sf(data = bee_sf_clean, aes(color = Plant_species), size = 2, alpha = 0.8) +
  labs(
    title = "Host Plant Observations in California",
    color = "Plant Species"
  ) +
  theme_minimal()

p_ca

ggsave(
  filename = file.path(output_dir, "host_plant_observations_california.png"),
  plot = p_ca,
  width = 10,
  height = 8,
  dpi = 300
)
```

## Statewide map for the top 10 host plants

```{r}
#| label: top-plants-statewide

top_plants <- bee_sf_clean %>%
  st_drop_geometry() %>%
  count(Plant_species, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(Plant_species)

bee_top <- bee_sf_clean %>%
  filter(Plant_species %in% top_plants)

p_top <- ggplot() +
  geom_sf(data = ca_counties, fill = "gray95", color = "gray50", linewidth = 0.2) +
  geom_sf(data = bee_top, aes(color = Plant_species), size = 2, alpha = 0.8) +
  labs(
    title = "Top Host Plant Observations in California",
    color = "Plant Species"
  ) +
  theme_minimal()

p_top

ggsave(
  filename = file.path(output_dir, "top_host_plant_observations_california.png"),
  plot = p_top,
  width = 10,
  height = 8,
  dpi = 300
)
```

## Session notes

- The original CRS mismatch error happened because `bee_sf` and `santa_barbara` were not in the same coordinate reference system.
- This script fixes that by transforming county geometries to `st_crs(bee_sf_clean)` before the spatial join.
- The earlier `st_as_sf()` error happened because `sf` had not yet been loaded.

