---
title: "Downloading Bee Records with Associated Host Plants from GBIF"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

## Overview

GBIF is primarily an occurrence database, not a curated bee--plant interaction database. Many useful bee--plant records are found in Darwin Core fields such as `associatedTaxa`, `associatedOrganisms`, `behavior`, and `occurrenceRemarks`.

This document shows one practical workflow for downloading bee records from GBIF and extracting records that may contain host plant, floral visitor, pollen, nectar, or other plant-association information.

## Setup

```{r}
# Install packages if needed
# install.packages(c("rgbif", "dplyr", "stringr", "readr", "purrr"))

library(rgbif)
library(dplyr)
library(stringr)
library(readr)
library(purrr)
```

## GBIF credentials

GBIF occurrence downloads require a GBIF account. Store your credentials in your `.Renviron` file rather than typing them directly into scripts.

```{r, eval=FALSE}
# Run this once to edit your .Renviron file:
# usethis::edit_r_environ()

# Add lines like these, replacing with your own credentials:
# GBIF_USER="lebuhn"
# GBIF_PWD="gbif4glb"
# GBIF_EMAIL="lebuhn@sfsu.edu"
```

Restart R after editing `.Renviron`.

## Define bee families

Here we use the major bee families. GBIF taxon keys are retrieved dynamically using the GBIF backbone taxonomy.

```{r}
library(rgbif)
library(dplyr)
library(stringr)
library(readr)
library(tidyr)

# Look up the GBIF backbone key for the bumble bee genus Bombus
bombus_key <- rgbif::name_backbone(
  name = "Bombus",
  rank = "GENUS",
  kingdom = "Animalia"
)$usageKey

bombus_key
```

```{r}
bombus_download <- occ_download(
  pred_default(),
  pred("genusKey", bombus_key),
  pred("country", "US"),
  pred("stateProvince", "California"),
  format = "DWCA"
)
```

```{r host-plant}

read_tsv_chr <- function(path) {
  if (!file.exists(path)) {
    return(tibble())
  }

  readr::read_tsv(
    path,
    col_types = cols(.default = col_character()),
    show_col_types = FALSE
  )
}

occurrence <- read_tsv_chr(file.path(out_dir, "occurrence.txt"))
verbatim   <- read_tsv_chr(file.path(out_dir, "verbatim.txt"))
```

```{r}
# Keep only records identified as Bombus crotchii

crotchii_occurrence <- occurrence %>%
  filter(
    species == "Bombus crotchii" |
      str_detect(scientificName, regex("^Bombus crotchii\\b", ignore_case = TRUE))
  )

dir.create("data", showWarnings = FALSE, recursive = TRUE)

crotchii_date <- format(Sys.Date(), "%Y-%m-%d")

crotchii_outfile <- file.path(
  "data",
  paste0("gbif_crotchii_", crotchii_date, ".csv")
)

readr::write_csv(crotchii_occurrence, crotchii_outfile)

crotchii_outfile
nrow(crotchii_occurrence)
getwd()
```
