---
title: "California Bombus Host-Plant Records from GBIF"
author: "Gretchen"
format: html
execute:
  echo: true
  warning: false
  message: true
params:
  gbif_download_key: ""
  refresh_checkpoints: false
---

## Overview

This document downloads California occurrence records for the bumble bee genus *Bombus* from GBIF and searches the resulting Darwin Core Archive for possible host-plant or floral-association information.

Host-plant information is not available as a searchable GBIF download predicate, so it cannot be used to limit the occurrence download directly. Here, we first download California occurrence records for the genus *Bombus* as a Darwin Core Archive, then search the downloaded occurrence, verbatim, and extension files for fields that may contain host-plant or floral-association information.

In a GBIF Darwin Core Archive, extension files are extra tables that hang off the main occurrence.txt table.

A useful way to think about the archive is:

File type Role Plain meaning occurrence.txt Main table One row per occurrence record, usually the bee specimen or observation verbatim.txt Original/source wording Less-interpreted text or fields from the publisher Extension files Related detail tables Extra information attached to occurrence records

Extension files are used when a record has information that does not fit neatly into the main occurrence table, or when there may be multiple related details per occurrence.

Because host-plant associations are often supplied as free text and are not standardized across all publishers, the extracted records should be treated as candidate host-plant associations requiring review before analysis.

## Project folder structure

This document is designed to live in the `scripts/` folder of a project with this structure:

``` text
project/
  scripts/
  data/
    raw/
    processed/
    gbif_bombus_california_dwca/
    gbif_downloads/
  outputs/
```

The script finds the project root automatically by looking upward for the `scripts/`, `data/`, and `outputs/` folders. This avoids hard-coded paths and should allow the project to run on any computer with the same folder structure.

GBIF zip files are stored in `data/gbif_downloads/`. If an older folder named `data/gbif download_files/` or `data/gibf download_files/` exists, the script checks those folders for existing zip files and copies them into `data/gbif_downloads/`.

## Checkpoint workflow

This workflow saves intermediate products as CSV files.

Raw downloaded or parsed data are saved in:

``` text
data/raw/
```

Processed intermediate files are saved in:

``` text
data/processed/
```

Final outputs are saved in:

``` text
outputs/
```

By default, existing checkpoint files are reused. To rebuild them from scratch, change this YAML parameter to `true`:

``` yaml
refresh_checkpoints: true
```

## Inputs and Outputs

| File | Role in workflow | What it is | Occurrence data? | Verbatim data? | Extension data? | Host-related filtering applied? | What changed from previous step? |
|----|----|----|----|----|----|----|----|
| `data/gbif_downloads/<gbif_download_key>.zip` | Input | Original GBIF Darwin Core Archive download | Yes | Maybe | Maybe | No | Original GBIF download; no R processing yet. |
| `data/raw/00_bombus_taxon_key.csv` | Reference/tracking output | GBIF backbone taxon-key lookup for *Bombus* | No | No | No | No | Created before the download; identifies the GBIF taxon key used for the search. |
| `data/raw/01_gbif_download_key.csv` | Reference/tracking output | GBIF download key record | No | No | No | No | Records which GBIF download was used or requested. |
| `data/raw/02_occurrence_raw.csv` | Intermediate output | Parsed copy of `occurrence.txt` | Yes | No | No | No | `occurrence.txt` was extracted from the zip and saved as CSV. |
| `data/raw/03_verbatim_raw.csv` | Intermediate output | Parsed copy of `verbatim.txt`, if present | No | Yes | No | No | `verbatim.txt` was extracted from the zip and saved as CSV; if absent, an empty placeholder table is created. |
| `data/processed/04_host_column_inventory.csv` | Intermediate output | Inventory of likely host-related columns in occurrence and verbatim files | Yes | Yes | No | Column-name screen only | Column names were screened for possible host/floral relevance. Individual record text was not filtered. |
| `data/processed/05_host_from_occurrence.csv` | Intermediate output | Candidate host records from `occurrence.txt` | Yes | No | No | Host-text filter | Likely host-related occurrence columns were searched for host/floral text terms. |
| `data/processed/06_host_from_verbatim.csv` | Intermediate output | Candidate host records from `verbatim.txt` | No | Yes | No | Host-text filter | Likely host-related verbatim columns were searched for host/floral text terms. |
| `data/processed/07_extension_file_inventory.csv` | Intermediate output | Inventory of extension files | No | No | Yes | No | The script identified extra `.txt` files in the archive, excluding `occurrence.txt`, `verbatim.txt`, `citations.txt`, and `rights.txt`. |
| `data/processed/08_host_from_extensions.csv` | Intermediate output | Candidate host records from extension files | No | No | Yes | Host-text filter | Likely host-related extension columns were searched for host/floral text terms. |
| `data/processed/09_candidate_host_records_all.csv` | Intermediate output | Combined candidate host records | Yes | Yes | Yes | Host-text filtered source | Candidate records from occurrence, verbatim, and extension files were combined, deduplicated, and sorted. |
| `data/processed/10_candidate_host_record_summary.csv` | Intermediate output | Summary of candidate host records | Yes | Yes | Yes | Host-text filtered source | Combined candidate records were summarized by source file and host field. |
| `outputs/gbif_bombus_california_candidate_host_records.csv` | Final output | Final exported candidate host-record CSV | Yes | Yes | Yes | Host-text filtered source | Export copy of `09_candidate_host_records_all.csv` for downstream review. |

## Setup

This chunk loads packages, finds the project root, and defines all project-relative paths. No absolute paths are used.

```{r setup}
# install.packages(c(
#   "rgbif",
#   "dplyr",
#   "stringr",
#   "readr",
#   "tidyr",
#   "purrr",
#   "tibble"
# ))

library(rgbif)
library(dplyr)
library(stringr)
library(readr)
library(tidyr)
library(purrr)
library(tibble)

has_text <- function(x) {
  length(x) == 1 && !is.na(x) && nzchar(x)
}

find_project_root <- function() {
  input_file <- tryCatch(knitr::current_input(), error = function(e) NULL)

  if (has_text(input_file) && file.exists(input_file)) {
    start_dir <- dirname(normalizePath(input_file, winslash = "/", mustWork = TRUE))
  } else {
    start_dir <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
  }

  current_dir <- start_dir

  for (i in seq_len(10)) {
    has_project_dirs <- all(dir.exists(file.path(
      current_dir,
      c("scripts", "data", "outputs")
    )))

    if (has_project_dirs) {
      return(current_dir)
    }

    parent_dir <- dirname(current_dir)

    if (identical(parent_dir, current_dir)) {
      break
    }

    current_dir <- parent_dir
  }

  stop(
    "Could not find project root. Expected to find scripts/, data/, and outputs/ ",
    "in the project root or one of its parent directories."
  )
}

project_root <- find_project_root()

path_from_root <- function(...) {
  file.path(project_root, ...)
}

scripts_dir <- path_from_root("scripts")
data_dir <- path_from_root("data")
raw_dir <- path_from_root("data", "raw")
processed_dir <- path_from_root("data", "processed")
dwca_dir <- path_from_root("data", "gbif_bombus_california_dwca")
gbif_download_dir <- path_from_root("data", "gbif_downloads")
outputs_dir <- path_from_root("outputs")

# Compatibility with older or misspelled folder names.
legacy_gbif_download_dirs <- c(
  path_from_root("data", "gbif download_files"),
  path_from_root("data", "gibf download_files")
)

walk(
  c(raw_dir, processed_dir, dwca_dir, gbif_download_dir, outputs_dir),
  dir.create,
  showWarnings = FALSE,
  recursive = TRUE
)

final_output_csv <- file.path(
  outputs_dir,
  "gbif_bombus_california_candidate_host_records.csv"
)

message("Project root: ", project_root)
message("GBIF zip files will be stored in: ", gbif_download_dir)
message("Final output will be written to: ", final_output_csv)
```

## GBIF credentials

GBIF downloads require a GBIF account. Store credentials in `.Renviron` rather than writing them directly into this document.

Your `.Renviron` file should contain lines like these, using your own GBIF account information:

```{r credentials-example, eval=FALSE}
GBIF_USER="your_gbif_username"
GBIF_PWD="your_gbif_password"
GBIF_EMAIL="your_email@example.edu"
```

After editing `.Renviron`, restart R so the credentials are available through `Sys.getenv()`.

## Helper functions

This chunk defines helper functions for reading, writing, and reusing checkpoint CSV files.

```{r helper-functions}
write_checkpoint <- function(dat, path) {
  dir.create(dirname(path), showWarnings = FALSE, recursive = TRUE)
  readr::write_csv(dat, path, na = "")
  invisible(path)
}

read_checkpoint <- function(path) {
  readr::read_csv(
    path,
    col_types = readr::cols(.default = readr::col_character()),
    show_col_types = FALSE
  )
}

cached_csv <- function(path, make_data, rebuild = params$refresh_checkpoints) {
  if (file.exists(path) && !isTRUE(rebuild)) {
    message("Reading checkpoint: ", path)
    return(read_checkpoint(path))
  }

  message("Creating checkpoint: ", path)
  dat <- make_data()

  if (!inherits(dat, "data.frame")) {
    dat <- tibble::as_tibble(dat)
  }

  write_checkpoint(dat, path)
  dat
}

empty_file_tbl <- function() {
  tibble(.empty_file = character())
}

read_tsv_chr <- function(path, required = FALSE) {
  if (!file.exists(path)) {
    if (isTRUE(required)) {
      stop("Required file not found: ", path)
    } else {
      return(empty_file_tbl())
    }
  }

  readr::read_tsv(
    path,
    col_types = readr::cols(.default = readr::col_character()),
    show_col_types = FALSE
  )
}

empty_host_tbl <- function() {
  tibble(
    source_file = character(),
    source_row = character(),
    host_field = character(),
    host_text = character()
  )
}
```

## Checkpoint index

This chunk writes a simple index of the checkpoint files created by the workflow.

```{r checkpoint-index}
checkpoint_index <- tibble(
  checkpoint_file = c(
    file.path("data", "raw", "00_bombus_taxon_key.csv"),
    file.path("data", "raw", "01_gbif_download_key.csv"),
    file.path("data", "raw", "02_occurrence_raw.csv"),
    file.path("data", "raw", "03_verbatim_raw.csv"),
    file.path("data", "processed", "04_host_column_inventory.csv"),
    file.path("data", "processed", "05_host_from_occurrence.csv"),
    file.path("data", "processed", "06_host_from_verbatim.csv"),
    file.path("data", "processed", "07_extension_file_inventory.csv"),
    file.path("data", "processed", "08_host_from_extensions.csv"),
    file.path("data", "processed", "09_candidate_host_records_all.csv"),
    file.path("data", "processed", "10_candidate_host_record_summary.csv"),
    file.path("outputs", "gbif_bombus_california_candidate_host_records.csv")
  ),
  purpose = c(
    "GBIF backbone taxon key for Bombus.",
    "GBIF download key used to retrieve the Darwin Core Archive.",
    "Raw occurrence.txt table saved as CSV.",
    "Raw verbatim.txt table saved as CSV, when present.",
    "Inventory of occurrence and verbatim columns that may contain host-plant information.",
    "Candidate host-plant records extracted from occurrence.txt.",
    "Candidate host-plant records extracted from verbatim.txt.",
    "Inventory of Darwin Core Archive extension files.",
    "Candidate host-plant records extracted from extension files.",
    "Combined candidate host-plant records from all sources.",
    "Summary counts by source file and host-text field.",
    "Final candidate host-record CSV for downstream review."
  )
)

write_checkpoint(
  checkpoint_index,
  file.path(processed_dir, "00_checkpoint_index.csv")
)

checkpoint_index
```

## Look up the GBIF taxon key for *Bombus*

GBIF downloads use numeric taxon keys. This chunk looks up the GBIF backbone key for the genus *Bombus* and saves it in `data/raw/`.

```{r bombus-key}
bombus_key_tbl <- cached_csv(
  file.path(raw_dir, "00_bombus_taxon_key.csv"),
  function() {
    bombus_backbone <- rgbif::name_backbone(
      name = "Bombus",
      rank = "GENUS",
      kingdom = "Animalia"
    )

    if (is.null(bombus_backbone$usageKey) || is.na(bombus_backbone$usageKey)) {
      stop("No GBIF backbone usageKey found for Bombus.")
    }

    tibble(
      scientific_name = "Bombus",
      rank = "GENUS",
      kingdom = "Animalia",
      usageKey = as.character(bombus_backbone$usageKey),
      lookup_date = as.character(Sys.Date())
    )
  }
)

bombus_key <- as.integer(bombus_key_tbl$usageKey[[1]])

bombus_key_tbl
```

## Submit or reuse the GBIF download

This chunk either reuses an existing GBIF download key or submits a new GBIF download request.

The order of preference is:

1.  Use `params$gbif_download_key` if it is supplied in the YAML.
2.  Otherwise, reuse `data/raw/01_gbif_download_key.csv` if it already exists.
3.  Otherwise, submit a new GBIF download request.

```{r gbif-download}
download_key_checkpoint <- file.path(raw_dir, "01_gbif_download_key.csv")

gbif_param_key <- params$gbif_download_key

if (is.null(gbif_param_key) || is.na(gbif_param_key)) {
  gbif_param_key <- ""
}

if (has_text(gbif_param_key)) {

  download_key_tbl <- tibble(
    download_key = gbif_param_key,
    source = "YAML parameter",
    created_or_reused_at = as.character(Sys.time())
  )

  write_checkpoint(download_key_tbl, download_key_checkpoint)

} else if (file.exists(download_key_checkpoint) && !isTRUE(params$refresh_checkpoints)) {

  download_key_tbl <- read_checkpoint(download_key_checkpoint)

} else {

  gbif_user <- Sys.getenv("GBIF_USER")
  gbif_pwd <- Sys.getenv("GBIF_PWD")
  gbif_email <- Sys.getenv("GBIF_EMAIL")

  missing_gbif_env <- c(
    GBIF_USER = gbif_user,
    GBIF_PWD = gbif_pwd,
    GBIF_EMAIL = gbif_email
  )

  missing_gbif_env <- names(missing_gbif_env)[!nzchar(missing_gbif_env)]

  if (length(missing_gbif_env) > 0) {
    stop(
      "Missing GBIF credentials in .Renviron: ",
      paste(missing_gbif_env, collapse = ", ")
    )
  }

  bombus_download <- rgbif::occ_download(
    rgbif::pred_default(),
    rgbif::pred("genusKey", bombus_key),
    rgbif::pred("country", "US"),
    rgbif::pred("stateProvince", "California"),
    format = "DWCA",
    user = gbif_user,
    pwd = gbif_pwd,
    email = gbif_email
  )

  bombus_download <- rgbif::occ_download_wait(bombus_download)

  download_key_tbl <- tibble(
    download_key = bombus_download$key,
    source = "new GBIF download request",
    created_or_reused_at = as.character(Sys.time())
  )

  write_checkpoint(download_key_tbl, download_key_checkpoint)
}

bombus_download_key <- download_key_tbl$download_key[[1]]

download_key_tbl
```

## Retrieve and unzip the Darwin Core Archive

This chunk downloads the GBIF zip file into `data/gbif_downloads/`, then unzips it into `data/gbif_bombus_california_dwca/`.

If an older GBIF download folder exists, the script checks for the needed zip file there and copies it into the canonical folder.

```{r retrieve-download}
raw_core_checkpoints_exist <- all(file.exists(c(
  file.path(raw_dir, "02_occurrence_raw.csv"),
  file.path(raw_dir, "03_verbatim_raw.csv")
)))

extension_checkpoints_exist <- all(file.exists(c(
  file.path(processed_dir, "07_extension_file_inventory.csv"),
  file.path(processed_dir, "08_host_from_extensions.csv")
)))

dwca_needed <- isTRUE(params$refresh_checkpoints) ||
  !raw_core_checkpoints_exist ||
  !extension_checkpoints_exist

zip_filename <- paste0(bombus_download_key, ".zip")
zip_path <- file.path(gbif_download_dir, zip_filename)

legacy_zip_paths <- file.path(legacy_gbif_download_dirs, zip_filename)
legacy_zip_paths <- legacy_zip_paths[file.exists(legacy_zip_paths)]

if (!file.exists(zip_path) && length(legacy_zip_paths) > 0) {
  message("Copying existing GBIF zip from older folder: ", legacy_zip_paths[[1]])
  file.copy(
    from = legacy_zip_paths[[1]],
    to = zip_path,
    overwrite = TRUE
  )
}

if (dwca_needed) {

  if (!file.exists(zip_path) || isTRUE(params$refresh_checkpoints)) {
    rgbif::occ_download_get(
      key = bombus_download_key,
      path = gbif_download_dir,
      overwrite = isTRUE(params$refresh_checkpoints)
    )
  }

  if (!file.exists(zip_path)) {
    stop("Expected GBIF zip file was not found: ", zip_path)
  }

  if (dir.exists(dwca_dir) && isTRUE(params$refresh_checkpoints)) {
    unlink(
      list.files(dwca_dir, full.names = TRUE),
      recursive = TRUE,
      force = TRUE
    )
  }

  dir.create(dwca_dir, showWarnings = FALSE, recursive = TRUE)

  unzip(
    zipfile = zip_path,
    exdir = dwca_dir,
    overwrite = TRUE
  )

} else {
  message("Raw and extension checkpoints already exist; skipping GBIF zip retrieval.")
}

dwca_files <- if (dir.exists(dwca_dir)) {
  list.files(dwca_dir)
} else {
  character()
}

tibble(dwca_file = dwca_files)
```

## Define host-plant search helpers

This chunk defines the column and text-search logic used to find possible host-plant or floral-association information.

```{r host-search-helpers}
find_host_columns <- function(x) {
  if (ncol(x) == 0) {
    return(character())
  }

  names(x)[
    stringr::str_detect(
      names(x),
      stringr::regex(
        paste(
          c(
            "associatedTaxa",
            "associatedOrganisms",
            "associatedOccurrences",
            "host",
            "plant",
            "flower",
            "floral",
            "nectar",
            "pollen",
            "food",
            "substrate",
            "habitat",
            "occurrenceRemarks",
            "fieldNotes",
            "dynamicProperties"
          ),
          collapse = "|"
        ),
        ignore_case = TRUE
      )
    )
  ]
}

host_text_pattern <- stringr::regex(
  paste(
    c(
      "\\bhost\\b",
      "\\bplant\\b",
      "\\bflower\\b",
      "\\bfloral\\b",
      "\\bnectar\\b",
      "\\bpollen\\b",
      "forag",
      "\\bvisit",
      "\\bon\\b",
      "collected on",
      "reared from"
    ),
    collapse = "|"
  ),
  ignore_case = TRUE
)

make_host_column_inventory <- function(dat, source_file) {
  host_cols <- find_host_columns(dat)

  if (length(host_cols) == 0) {
    return(tibble(
      source_file = character(),
      host_field = character(),
      n_nonempty = character(),
      example = character()
    ))
  }

  purrr::map_dfr(
    host_cols,
    function(col) {
      vals <- dat[[col]]
      vals <- vals[!is.na(vals) & vals != ""]

      tibble(
        source_file = source_file,
        host_field = col,
        n_nonempty = as.character(length(vals)),
        example = if (length(vals) > 0) vals[[1]] else NA_character_
      )
    }
  )
}
```

## Read occurrence and verbatim files

This chunk reads the core Darwin Core files and saves them as CSV files in `data/raw/`.

```{r read-core-files}
occurrence <- cached_csv(
  file.path(raw_dir, "02_occurrence_raw.csv"),
  function() {
    read_tsv_chr(
      file.path(dwca_dir, "occurrence.txt"),
      required = TRUE
    )
  }
)

verbatim <- cached_csv(
  file.path(raw_dir, "03_verbatim_raw.csv"),
  function() {
    read_tsv_chr(
      file.path(dwca_dir, "verbatim.txt"),
      required = FALSE
    )
  }
)

core_file_counts <- tibble(
  source_file = c("occurrence.txt", "verbatim.txt"),
  n_rows = c(nrow(occurrence), nrow(verbatim)),
  n_columns = c(ncol(occurrence), ncol(verbatim))
)

core_file_counts
```

## Inspect likely host-plant columns

This chunk identifies columns in `occurrence.txt` and `verbatim.txt` that may contain host-plant, floral visitation, habitat, substrate, or related natural-history information.

```{r inspect-host-columns}
host_column_inventory <- cached_csv(
  file.path(processed_dir, "04_host_column_inventory.csv"),
  function() {
    bind_rows(
      make_host_column_inventory(occurrence, "occurrence.txt"),
      make_host_column_inventory(verbatim, "verbatim.txt")
    )
  }
)

host_column_inventory
```

## Extract possible host-plant text from occurrence and verbatim files

This chunk defines a function that pivots likely host-plant fields into long format and keeps rows where the field contains host-plant or floral-association search terms.

```{r extract-core-function}
extract_host_text <- function(dat, source_file) {
  if (nrow(dat) == 0) {
    return(empty_host_tbl())
  }

  host_cols <- find_host_columns(dat)

  if (length(host_cols) == 0) {
    return(empty_host_tbl())
  }

  id_cols <- intersect(
    c(
      "gbifID",
      "occurrenceID",
      "catalogNumber",
      "recordNumber",
      "datasetKey",
      "scientificName",
      "family",
      "genus",
      "specificEpithet",
      "eventDate",
      "decimalLatitude",
      "decimalLongitude",
      "locality",
      "county",
      "stateProvince",
      "countryCode",
      "basisOfRecord",
      "institutionCode",
      "collectionCode"
    ),
    names(dat)
  )

  dat %>%
    mutate(source_row = as.character(row_number())) %>%
    select(any_of(c("source_row", id_cols, host_cols))) %>%
    pivot_longer(
      cols = any_of(host_cols),
      names_to = "host_field",
      values_to = "host_text"
    ) %>%
    mutate(source_file = source_file, .before = source_row) %>%
    filter(
      !is.na(host_text),
      host_text != "",
      str_detect(host_text, host_text_pattern)
    )
}
```

## Save candidate records from occurrence.txt

This chunk extracts candidate host-plant records from `occurrence.txt` and saves them in `data/processed/`.

```{r extract-occurrence-host-records}
host_from_occurrence <- cached_csv(
  file.path(processed_dir, "05_host_from_occurrence.csv"),
  function() {
    extract_host_text(occurrence, "occurrence.txt")
  }
)

host_from_occurrence
```

## Save candidate records from verbatim.txt

This chunk extracts candidate host-plant records from `verbatim.txt` and saves them in `data/processed/`.

```{r extract-verbatim-host-records}
host_from_verbatim <- cached_csv(
  file.path(processed_dir, "06_host_from_verbatim.csv"),
  function() {
    extract_host_text(verbatim, "verbatim.txt")
  }
)

host_from_verbatim
```

## Search extension files

Darwin Core Archives may include extension files with measurements, facts, organism interactions, or relationships among records. This chunk inventories those files and saves the inventory in `data/processed/`.

```{r extension-files}
extension_file_inventory <- cached_csv(
  file.path(processed_dir, "07_extension_file_inventory.csv"),
  function() {
    all_txt <- list.files(
      dwca_dir,
      pattern = "\\.txt$",
      full.names = TRUE
    )

    extension_files <- all_txt[
      !basename(all_txt) %in% c(
        "occurrence.txt",
        "verbatim.txt",
        "citations.txt",
        "rights.txt"
      )
    ]

    tibble(
      source_file = basename(extension_files),
      path = extension_files,
      size_bytes = as.character(file.size(extension_files))
    )
  }
)

extension_files <- extension_file_inventory$path

extension_file_inventory
```

## Extract possible host-plant text from extension files

This chunk defines a function for extracting possible host-plant or floral-association information from Darwin Core Archive extension files.

```{r extract-extension-function}
extract_from_extension <- function(path) {
  dat <- read_tsv_chr(path)

  if (nrow(dat) == 0) {
    return(empty_host_tbl())
  }

  possible_cols <- names(dat)[
    str_detect(
      names(dat),
      regex(
        paste(
          c(
            "measurementType",
            "verbatimMeasurementType",
            "measurementValue",
            "measurementRemarks",
            "relationship",
            "resource",
            "host",
            "plant",
            "flower",
            "nectar",
            "pollen",
            "associated"
          ),
          collapse = "|"
        ),
        ignore_case = TRUE
      )
    )
  ]

  if (length(possible_cols) == 0) {
    return(empty_host_tbl())
  }

  id_cols <- intersect(
    c(
      "gbifID",
      "id",
      "coreid",
      "coreId",
      "occurrenceID",
      "resourceID",
      "measurementID",
      "relationshipOfResourceID"
    ),
    names(dat)
  )

  dat %>%
    mutate(source_row = as.character(row_number())) %>%
    select(any_of(c("source_row", id_cols, possible_cols))) %>%
    pivot_longer(
      cols = any_of(possible_cols),
      names_to = "host_field",
      values_to = "host_text"
    ) %>%
    mutate(source_file = basename(path), .before = source_row) %>%
    filter(
      !is.na(host_text),
      host_text != "",
      str_detect(host_text, host_text_pattern)
    )
}
```

```{r extract-extension-host-records}
host_from_extensions <- cached_csv(
  file.path(processed_dir, "08_host_from_extensions.csv"),
  function() {
    if (length(extension_files) == 0) {
      return(empty_host_tbl())
    }

    purrr::map_dfr(extension_files, extract_from_extension)
  }
)

host_from_extensions
```

## Combine candidate host-plant records

This chunk combines candidate host records from occurrence, verbatim, and extension files. The combined table is saved in `data/processed/` and also written to `outputs/` as the main final CSV.

```{r combine-host-records}
host_records_all <- bind_rows(
  host_from_occurrence,
  host_from_verbatim,
  host_from_extensions
) %>%
  distinct() %>%
  arrange(source_file, source_row, host_field)

write_checkpoint(
  host_records_all,
  file.path(processed_dir, "09_candidate_host_records_all.csv")
)

readr::write_csv(
  host_records_all,
  final_output_csv,
  na = ""
)

host_records_all
```

## Summarize extracted records

This chunk summarizes the number of candidate host-plant records by source file and host-text field.

```{r summarize-host-records}
host_record_summary <- host_records_all %>%
  count(source_file, host_field, sort = TRUE)

write_checkpoint(
  host_record_summary,
  file.path(processed_dir, "10_candidate_host_record_summary.csv")
)

host_record_summary
```

## Output files

This chunk prints the main output locations so they are visible in the rendered HTML file.

```{r output-files}
tibble(
  output_type = c(
    "Project root",
    "Raw data directory",
    "Processed data directory",
    "GBIF zip directory",
    "Unzipped Darwin Core Archive directory",
    "Outputs directory",
    "Final candidate host-record CSV"
  ),
  path = c(
    project_root,
    raw_dir,
    processed_dir,
    gbif_download_dir,
    dwca_dir,
    outputs_dir,
    final_output_csv
  )
)
```

## Notes for downstream cleaning

The resulting CSV should be reviewed before analysis. In particular:

- Some records may describe adult floral visitation rather than larval host use.
- Some plant names may be informal, misspelled, abbreviated, or embedded in longer notes.
- Some records may mention habitat or substrate rather than an actual plant association.
- Duplicate host information may appear in both interpreted and verbatim files.
- Records lacking host-plant information are expected and are not errors.
- The text search used here is intentionally broad, so false positives should be expected.
