---
title: "California Bombus Host-Plant Records from GBIF"
author: "Gretchen"
format: html
execute:
  echo: true
  warning: false
  message: true
params:
  gbif_download_key: ""
  download_dir: "gbif_downloads"
  out_dir: "gbif_bombus_california_dwca"
  output_csv: "gbif_bombus_california_candidate_host_records.csv"
---

## Overview

This script downloads California occurrence records for the bumble bee genus *Bombus* from GBIF and searches the resulting Darwin Core Archive for possible host-plant or floral-association information.

Host-plant information is not available as a searchable GBIF download predicate, so it cannot be used to limit the occurrence download directly. Here, we first download California occurrence records for the genus *Bombus* as a Darwin Core Archive, then search the downloaded occurrence, verbatim, and extension files for fields that may contain host-plant or floral-association information.

Because host-plant associations are often supplied as free text and are not standardized across all publishers, the extracted records should be treated as candidate host-plant associations requiring review before analysis.

## Setup

```{r setup}
# install.packages(c(
#   "rgbif",
#   "dplyr",
#   "stringr",
#   "readr",
#   "tidyr",
#   "purrr"
# ))

library(rgbif)
library(dplyr)
library(stringr)
library(readr)
library(tidyr)
library(purrr)
```

## GBIF credentials

GBIF downloads require a GBIF account. The cleanest approach is to store your credentials in `.Renviron` rather than writing them directly into this script.

Your `.Renviron` file should contain lines like these, using your own GBIF account information:

```{r credentials-example, eval=FALSE}
GBIF_USER="your_gbif_username"
GBIF_PWD="your_gbif_password"
GBIF_EMAIL="your_email@example.edu"
```


## Look up the GBIF taxon key for *Bombus*

```{r bombus-key}
bombus_key <- rgbif::name_backbone(
  name = "Bombus",
  rank = "GENUS",
  kingdom = "Animalia"
)$usageKey

bombus_key
```

## Download California *Bombus* occurrence records

This download uses `pred_default()`, which keeps records that are present, have coordinates, and lack flagged geospatial issues. The additional predicates restrict the download to the genus *Bombus*, the United States, and California.

If you already submitted a GBIF download and have its download key, enter it in the YAML parameter `gbif_download_key` at the top of this document. If `params$gbif_download_key` is blank, the script will submit a new GBIF download.

```{r gbif-download}
if (nzchar(params$gbif_download_key)) {
  bombus_download_key <- params$gbif_download_key
} else {
  bombus_download <- occ_download(
    pred_default(),
    pred("genusKey", bombus_key),
    pred("country", "US"),
    pred("stateProvince", "California"),
    format = "DWCA"
  )

  bombus_download <- occ_download_wait(bombus_download)
  bombus_download_key <- bombus_download$key
}

bombus_download_key

```

## Retrieve and unzip the Darwin Core Archive

```{r retrieve-download}
dir.create(params$out_dir, showWarnings = FALSE, recursive = TRUE)

zip_path <- file.path(
  params$download_dir,
  paste0(bombus_download_key, ".zip")
)

unzip(zip_path, exdir = params$out_dir)

list.files(params$out_dir)
```

## Helper functions

```{r helper-functions}
read_tsv_chr <- function(path) {
  if (!file.exists(path)) {
    return(tibble())
  }

  readr::read_tsv(
    path,
    col_types = cols(.default = col_character()),
    show_col_types = FALSE
  )
}

find_host_columns <- function(x) {
  if (ncol(x) == 0) {
    return(character())
  }

  names(x)[
    str_detect(
      names(x),
      regex(
        paste(
          c(
            "associatedTaxa",
            "associatedOrganisms",
            "associatedOccurrences",
            "host",
            "plant",
            "flower",
            "floral",
            "nectar",
            "pollen",
            "food",
            "substrate",
            "habitat",
            "occurrenceRemarks",
            "fieldNotes",
            "dynamicProperties"
          ),
          collapse = "|"
        ),
        ignore_case = TRUE
      )
    )
  ]
}

host_text_pattern <- regex(
  paste(
    c(
      "host",
      "plant",
      "flower",
      "floral",
      "nectar",
      "pollen",
      "forag",
      "visit",
      "on ",
      "collected on",
      "reared from"
    ),
    collapse = "|"
  ),
  ignore_case = TRUE
)
```

## Read occurrence and verbatim files

```{r read-core-files}
occurrence <- read_tsv_chr(file.path(params$out_dir, "occurrence.txt"))
verbatim   <- read_tsv_chr(file.path(params$out_dir, "verbatim.txt"))

nrow(occurrence)
nrow(verbatim)
```

## Inspect likely host-plant columns

```{r inspect-host-columns}
find_host_columns(occurrence)
find_host_columns(verbatim)
```

## Extract possible host-plant text from occurrence and verbatim files

```{r extract-core-host-records}
extract_host_text <- function(dat, source_file) {
  if (nrow(dat) == 0) {
    return(tibble())
  }

  host_cols <- find_host_columns(dat)

  if (length(host_cols) == 0) {
    return(tibble())
  }

  id_cols <- intersect(
    c(
      "gbifID",
      "occurrenceID",
      "catalogNumber",
      "recordNumber",
      "datasetKey",
      "scientificName",
      "family",
      "genus",
      "specificEpithet",
      "eventDate",
      "decimalLatitude",
      "decimalLongitude"
    ),
    names(dat)
  )

  dat %>%
    select(any_of(c(id_cols, host_cols))) %>%
    pivot_longer(
      cols = any_of(host_cols),
      names_to = "host_field",
      values_to = "host_text"
    ) %>%
    mutate(source_file = source_file) %>%
    filter(
      !is.na(host_text),
      host_text != "",
      str_detect(host_text, host_text_pattern)
    )
}

host_from_occurrence <- extract_host_text(occurrence, "occurrence.txt")
host_from_verbatim   <- extract_host_text(verbatim, "verbatim.txt")
```

## Search extension files

Darwin Core Archives may include extension files that contain additional information, such as measurements, facts, or relationships among organisms. These files are not guaranteed to be present, but when they are present they may contain host-plant information.

```{r extension-files}
all_txt <- list.files(params$out_dir, pattern = "\\.txt$", full.names = TRUE)

extension_files <- all_txt[
  !basename(all_txt) %in% c(
    "occurrence.txt",
    "verbatim.txt",
    "citations.txt",
    "rights.txt"
  )
]

basename(extension_files)
```

```{r extract-extension-host-records}
extract_from_extension <- function(path) {
  dat <- read_tsv_chr(path)

  if (nrow(dat) == 0) {
    return(tibble())
  }

  possible_cols <- names(dat)[
    str_detect(
      names(dat),
      regex(
        paste(
          c(
            "measurementType",
            "verbatimMeasurementType",
            "measurementValue",
            "measurementRemarks",
            "relationship",
            "resource",
            "host",
            "plant",
            "flower",
            "nectar",
            "pollen",
            "associated"
          ),
          collapse = "|"
        ),
        ignore_case = TRUE
      )
    )
  ]

  if (length(possible_cols) == 0) {
    return(tibble())
  }

  id_cols <- intersect(
    c(
      "gbifID",
      "id",
      "coreid",
      "coreId",
      "occurrenceID",
      "resourceID"
    ),
    names(dat)
  )

  dat %>%
    select(any_of(c(id_cols, possible_cols))) %>%
    pivot_longer(
      cols = any_of(possible_cols),
      names_to = "host_field",
      values_to = "host_text"
    ) %>%
    mutate(source_file = basename(path)) %>%
    filter(
      !is.na(host_text),
      host_text != "",
      str_detect(host_text, host_text_pattern)
    )
}

host_from_extensions <- purrr::map_dfr(extension_files, extract_from_extension)
```

## Combine candidate host-plant records

```{r combine-host-records}
host_records_all <- bind_rows(
  host_from_occurrence,
  host_from_verbatim,
  host_from_extensions
) %>%
  distinct()

host_records_all
```

## Summarize extracted records

```{r summarize-host-records}
host_records_all %>%
  count(source_file, host_field, sort = TRUE)
```

## Save candidate host-plant records

```{r save-results}
readr::write_csv(host_records_all, params$output_csv)

params$output_csv
```

## Notes for downstream cleaning

The resulting CSV should be reviewed before analysis. In particular:

- Some records may describe floral visitation rather than true larval host use.
- Some plant names may be informal, misspelled, abbreviated, or embedded in longer notes.
- Some records may mention habitat or substrate rather than an actual plant association.
- Duplicate host information may appear in both interpreted and verbatim files.
- Records lacking host-plant information are expected and are not errors.
